%% Copyright 2022 Yuanhao Chen
%
% This work may be distributed and/or modified under the
% conditions of the LaTeX Project Public License, either version 1.3
% of this license or (at your option) any later version.
% The latest version of this license is in
%   http://www.latex-project.org/lppl.txt
% and version 1.3 or later is part of all distributions of LaTeX
% version 2005/12/01 or later.
%
% This work has the LPPL maintenance status `maintained'.
% 
% The Current Maintainer of this work is Yuanhao Chen.
%
% This work consists of the files kanbun.sty, kanbun.lua,
% kanbun-luatex.sty, kanbun-example.tex and kanbun.tex.
% 
\documentclass{ltjtarticle}

\usepackage[match]{luatexja-fontspec}
\setmainjfont{FOT-TsukuBOldMin Pr6N}

\newlength\myzw
\setlength\myzw{2.25\zw}
\usepackage[kumi=beta, unit=\myzw, yokoaki=3.5]{kanbun}

\usepackage[margin=0pt]{geometry}

\usepackage{xcolor}

\Kanbun
初メテ，帝ノ母劉貴人ハ賜(たま)ハリ[レ]死ヲ，太祖ハ告ゲ[レ]帝ニ曰ハク「昔漢武帝ハ將‹し›«テ»ニ[レ]立タセント[二]其ノ子ヲ[一]而殺シ[二]其ノ母ヲ[一]，不[下]令(し)メ[三]婦人ヲ後ロニ與(あづか)ラ[二]國政ニ[一]，使(し)メ[中]外家ヲ為(な)サ[上レ]亂ヲ。汝當ニ‹べ›«ク»[二]繼統ス[一]，故ニ吾レ遠クシテ同ズルコト[二]漢武ト[一]，為(た)リ[二]長久之計[一]。」帝素ヨリ純孝ナリ，哀シク泣キ不[レ]能ハ[二]自ラ勝ツコト[一]，太祖怒ル[レ]之レヲ。帝ハ還リ[レ]宮ニ，哀シク不[二]自ラ止マ[一]，日夜ニシテ號泣ス。太祖知リテ而又タ召ス[レ]之レヲ。帝欲シ[レ]入ラント，左右曰ハク「孝子事ヘバ[レ]父ニ，小杖則チ受ケ，大杖避ク[レ]之レヲ。今陛下ノ怒ルコト盛ンナリ，入レバ或イハ不測ナリ，陷ラス[二]帝ヲ於不義ニ[一]。不[レ]如カ[二]且(しばら)ク出ヅルニ[一]，待チテ[二]怒リノ解クルヲ[一]而進ムコト，不ル[レ]晚カラ也。」帝ハ懼レ，從ヒ[レ]之レニ，乃チ游行シ逃グ[二]於外ニ[一]。
\EndKanbun

\begin{document}
\color[HTML]{e0e0e0}
\pagestyle{empty}
\vspace*{\fill}
\begin{minipage}{0.5\pageheight}
    \printkanbun
\end{minipage}
\vspace*{\fill}
\end{document}